\bf The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-10-18 paper.tex
changed format to jasa and edited reference so that it would work

---
title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
author: 
  - name: Harriet Mason
    url: https://harrietmason.netlify.app/
    orcid: 0009-0007-4568-8215
    email:  harriet.m.mason@gmail.com
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Dianne Cook
    url: https://dicook.org
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Sarah Goodwin
    url: https://www.linkedin.com/in/smgoodwin/
    orcid: 0000-0001-8894-8282
    email: Sarah.Goodwin@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Human Centred Computing
      city: Melbourne
      country: Australia
  - name: Emi Tanaka
    url: https://emitanaka.org/
    orcid: 0000-0002-1455-259X
    email: Emi.Tanaka@anu.edu.au
    affiliation: 
    - name: The Australian National University
      department: Biological Data Science Institute
      city: Canberra
      country: Australia
  - name: Susan Vanderplus
    url: https://srvanderplas.github.io
    orcid: 0000-0002-3803-0972
    email: susan.vanderplas@unl.edu
    affiliation:
    - name: University of Nebraska–Lincoln
      department: Statistics Department
      city:  Lincoln
      country: United States
bibliography: references.bib
abstract: Uncertainty visualisation is quickly becomming a hot topic in information visualisation. Exisiting reviews in the field take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information. This conflict largely stems from a conflation between uncertainty visualisations designed for decision making and those designed to prevent false conclusions. We coin the term "signal suppression" to describe a visualisation that is designed for preventing false conclusions, as the approach demands that the signal (i.e. the collective take away of the estimates) is suppressed by the noise (i.e. the variance on those estimates). We argue that the current standards in visualisation suggest that uncertainty visualisations designed for decision making should not be considered uncertainty visualisations at all. Therefore, future work should focus on signal suppression. Effective signal suppression requires us to communicate the signal and the noise as a single "validity of signal" variable, and doing so proves to be difficult with current methods. We illustrate current approaches to uncertainty visualisation by showing how they would change the visual apprearance of a choropleth map. These maps allow us to see why some methods succeed at signal suppression, while others fall short. Evaluating visualisations on how well they perform signal suppression also proves to be difficult, as it involves measuring the effect of noise, a variable we typically try to ignore. We suggest authors use qualitative studies or compare uncertainty visualisations to the relevant hypothesis tests. 
date: last-modified
toc: false
number-sections: true
format:
  jasa-pdf:
    keep-tex: true  
    journal:
      blinded: false
  jasa-html: default
fig-valign: bottom
cap-location: bottom
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| message: false
#| warning: false

# load Libraries
library(tidyverse)
# devtools::install_github("lydialucchesi/Vizumap")
library(Vizumap)
library(RColorBrewer)
library(scales)
library(sf)
library(ggrepel)
# devtools::install_github("UrbanInstitute/urbnmapr")
library(urbnmapr)
library(flextable)
library(colorspace)
library(rgeos)
```

## Introduction

@@

Among these summary tools, information visualisations are one of the most powerful as they allow for quick and memorable communication that allow us to identify quirks in our data that we didn't know to look for.
Datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] show a case where visual statistics highlight elements of the data that are invisible to the typical summary statistics. 
Something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].

"Uncertainty visualisation" is a relatively new field in research. 
Early mentions of uncertainty visualisation start to appear in the late 1980s [@Ibrekk1987], with geospatial information visualisation literature from the early 1990s declaring this to be essential aspect of any information display [@MacEachren1992; @Carr1992]. 
@fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers. 
Despite kicking off the field, these papers did not define uncertainty visualisation.
This has led to a lack of consensus on what it means for a graphic to visualise uncertainty, an issue we will return to later.
Therefore, while the field is considered to be quite new, many of the graphics used for uncertainty visualisation have been around for much longer. 
For example, box plots and histograms display variation which becomes synonymous with uncertainty when we are using them to depict the variation of an estimate. 
Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field.
That is, now that there is an overwhelming amount of information, it is valuable to distil it into simple facts. 
In fact, there have already been several reviews published but a central piece of discussion is missing.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ibrekk
#| fig-cap: "A replication of the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments. Several visualisation methods that are now unpopular (such as the pie chart) are used throughout this paper."
#| fig-subcap: 
#|   - "Picture 1"
#|   - "Picture 2"
#|   - "Picture 3"
#|   - "Picture 4"
#|   - "Picture 5"
#|   - "Picture 6"
#|   - "Picture 7"
#|   - "Picture 8"
#|   - "Picture 9"
#| layout-ncol: 3

# Generate data
set.seed(1)
x=rnorm(1000, 8, 4)
ib_data <- tibble(x=ifelse(x<0, -x, x))

# Picture 1
p1 <- ib_data |>
  summarise(avg = mean(x),
          conf_95a = quantile(x, probs=c(0.025)),
          conf_95b = quantile(x, probs=c(0.975))) |>
  ggplot(aes(y="NA")) +
  geom_point(aes(x=avg)) +
  geom_errorbar(aes(xmin = conf_95a, xmax = conf_95b), width = 0.1) +
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks=seq(0,19),
                     labels= ggplot2:::interleave(as.character(c(seq(0,18, 2), 19)), rep("", 11))[c(0:19, 21)],
                     limits=c(0,19)) +
  theme_classic() +
  theme(axis.line.y=element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        aspect.ratio=1/10)

# Picture 2
p2 <- ib_data |>
  mutate(x = ifelse(x>18, 18, x),
         binx = cut(x, breaks=seq(0,18,2))) |>
  group_by(binx) |>
  summarise(n = n()) |>
  mutate(Probability = n / sum(n)) |>
  ggplot(aes(x=binx, y=Probability)) +
  geom_col(fill="black", colour="white") +
  scale_x_discrete(name = "INCHES OF SNOW",
                   labels= paste0(seq(0,16,2), sep = "-", seq(2,18,2))) +
  scale_y_continuous(breaks = seq(0.00, 0.25, 0.05)) +
  theme_classic() + 
  theme(aspect.ratio=0.33)

# Picture 3
p3 <- ib_data |>
  mutate(x = ifelse(x>18, 18, x),
         binx = cut(x, 
                     breaks=seq(0,18,2), 
                     labels= paste0(seq(0,16,2), sep = "-", seq(2,18,2)))) |>
  group_by(binx) |>
  summarise(n = n()) |>
  mutate(Probability = n / sum(n),
         csum = rev(cumsum(rev(Probability))), 
         pos = Probability/2 + lead(csum, 1),
         pos = if_else(is.na(pos), Probability/2, pos)) |>
  ggplot(aes(x="", y=Probability, fill=binx)) +
  geom_bar(stat="identity", width=1) +
  geom_text_repel(aes(y = pos, label = paste0(round(Probability*100), sep="", "%")),
                   size = 3, nudge_x = 0.6, show.legend = FALSE, segment.color = 'transparent') +
  #geom_label(aes(label = paste0(round(Probability*100), sep="", "%")),
  #          position = position_stack(vjust = 0.5)) + 
  scale_fill_grey() +
  coord_polar("y", start=0) +
  labs(fill = "INCHES OF SNOW") + 
  theme_void() + 
  theme(aspect.ratio=1)

# Picture 4
p4 <- ib_data |>
  ggplot(aes(x=x)) +
  geom_density() + 
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks = seq(0,20,2),
                     labels= paste0(seq(0,20,2))) +
  scale_y_continuous(name = "Probability density",
                     breaks = seq(0.00, 0.20, 0.02)) +
  theme_classic() + 
  theme(aspect.ratio=0.33)

# Picture 5
p5 <- ib_data |>
  ggplot(aes(y="", x=x)) +
  geom_violin() + 
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks = seq(0,20,2),
                     labels= paste0(seq(0,20,2)),
                     limits=c(0,20)) +
  theme_classic() + 
  theme(axis.line.y=element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        aspect.ratio=0.4)

# Picture 6
set.seed(1)
x=rnorm(5000, 8, 4)
ib_data2 <- tibble(x=ifelse(x<0, -x, x)) |>
  mutate(x=ifelse(x>=18, 18-rexp(5000,rate=0), x))
p6 <- ib_data2 |>
  ggplot(aes(y="", x=x)) +
  geom_jitter(size=0.05) + 
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks = seq(0,20,2),
                     labels= paste0(seq(0,20,2)),
                     limits=c(0,20)) +
  theme_classic() + 
  theme(axis.line.y=element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        aspect.ratio=0.1)

# Picture 7
p7 <- ib_data2 |>
  arrange(x) |>
  mutate(group = rep(1:50, each=100))|>
  group_by(group) |>
  summarise(x = max(x, na.rm=TRUE)) |>
  add_row(group=c(0,51), x = c(0,20)) |>
  ggplot(aes(x=x)) +
  geom_linerange(ymin = 0.1, ymax = 1) + 
  geom_linerange(y=1, xmin = -0.03, xmax = 20.03)+ 
  geom_linerange(y=0.1, xmin = -0.03, xmax = 20.03)+ 
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks = seq(0,20,2),
                     labels= paste0(seq(0,20,2)),
                     limits=c(0,20)) +
  scale_y_continuous(limits=c(0,1)) + 
  theme_classic() + 
  theme(axis.line.y=element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        aspect.ratio=0.1)

# Picture 8
p8 <- ib_data |>
  reframe(x = quantile(x, probs=c(0.25, 0.50, 0.75)))|>
  add_row(x = c(0,20)) |>
  arrange(x) |>
  mutate(quantile = c("min", "q1", "med", "q3", "max")) |>
  pivot_wider(names_from = quantile, values_from = x) |>
  ggplot(aes(y="")) +
  #geom_point(aes(x=med)) +
  geom_errorbar(aes(y="", xmin = min, xmax = max), width = 0.2) +
  geom_crossbar(aes(y="", x=med, xmin = q1, xmax = q3), width = 0.5) +
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks=seq(0,20),
                     labels= ggplot2:::interleave(as.character(c(seq(0,20, 2))), rep("", 11))[1:21],
                     limits=c(0,20)) +
  theme_classic() +
  theme(axis.line.y=element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        aspect.ratio=1/10)

# Picture 9
p9 <- ib_data |>
  ggplot(aes(x)) +
  stat_ecdf(geom = "step") +
  scale_x_continuous(name = "INCHES OF SNOW",
                     breaks=seq(0,20),
                     labels= ggplot2:::interleave(as.character(c(seq(0,20, 2))), rep("", 11))[1:21],
                     limits=c(0,20)) +
  scale_y_continuous(name = "Cumulative probability",
                     breaks=seq(0,1,0.1),
                     labels= seq(0,1,0.1),
                     limits=c(0,1)) +
  theme_classic() +
  theme(aspect.ratio=4/10)

# Display Plots
p1
p2
p3
p4
p5
p6
p7
p8
p9
```

Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, but rather, they comment on the *difficulties* faced when trying to summarise the field. 
@Kinkeldey2014 found most experimental methods to be ad hoc, with no commonly agreed upon methodology, formalisations, or greater goal of describing general principals. 
@Hullman2016 noticed there is a serious noise issue in the data coming from uncertainty visualisation experiments. 
She commented on the prevalence of confounding variables that make it unclear as to what exactly caused a subjects poor performance on a set of particular questions.
Mistakes due to misunderstanding visualisations, misinterpreting questions, and incorrectly applying heuristics are all combined into a single error value.
@Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal best plot for all people and circumstances.
@Griethe2006 did not identify common themes, but instead listed the findings and opinions of a collection of papers. 
@uncertchap2022 summarised several cognitive effects that repeatedly arise in uncertainty visualisation experiments, however these effects were each discussed in isolation as a list of considerations an author might make rather than an overarching theory of rules for effective uncertainty visualisation. 
While these reviews are thorough in scope, none discuss how the existing literature contribute to the broader goal of uncertainty visualisation. 
The problem faced by the literature is easily summarised with a famous quote by Henri Poincaré.

> "Science is built up of facts, as a house is built of stones; but an accumulation of facts is no more a science than a heap of stones is a house." - Henri Poincaré (1905)

@@

This review is broken into several parts that each reflect a different approach to uncertainty visualisation. 
First, we look at graphics that ignore uncertainty entirely and discuss why uncertainty should be included at all. 
Second, we look at methods that consider uncertainty to be just another variable and discuss the characteristics of uncertainty that make it a unique visualisation problem. 
Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss the limitations of these approaches. 
Fourth, we discuss methods that implicitly include uncertainty by depicting a sample in place of an estimate. 
Finally, we discuss how uncertainty visualisations can be effectively evaluated.
When discussing each of these methods, we will repeatedly return to the *purpose* of uncertainty visualisation and the effectiveness of each approach in fulfilling that purpose.

### Spatial example
There are far too many uncertainty visualisations to exhaustively discuss them all. 
Instead we focus on the changes made to a single plot, the choropleth map. 
Due to the field's origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty. 
Utilising a single example will help isolate the ideas we are trying to convey. 
Therefore, it is important to remember that even though we focus our discussion on the choropleth map, the theoretical approach we outline in this review is useful to all uncertainty visualisations regardless of their application. 
Additionally, our examples focus on incorporating uncertainty through colour manipulation, as that is the key visual channel used in a choropleth map.
However, the methods we discuss go beyond variations in a colour palette.
Even though they are not explicitly shown, visualisations that depict uncertainty using layers such as position or shape, and more complicated graphics that incorporate animation or interactivity are also within the scope of this review.
We will use the choropleth maps as a tool to clearly highlight the costs and benefits of each approach.

@fig-data shows the first six rows and the geographical boundaries of our data set. 
The temperature variable was generated using a sine wave, that is $Temperature_i = 29-2\cdot{|Latitude_i - \sin{2 \cdot Longitude_i}|}$ where the $Longitude$ and $Latitude$ are the longitude and latitude of the county's centroid scaled to a standard normal distribution. 
Each county's variance is independently, randomly sampled from a uniform distribution. 
The low standard errors are drawn from a $U_{[0,1]}$ distribution, while the high standard errors are drawn from a $U_{[1,2]}$ distribution. 
As we are dealing with an average, the sampling distribution would be approximately normal by the central limit theorem, so each county temperature estimate is assumed to come from a $N(Temp_i, SE_{case,i})$. 
This is the data we will be using in our spatial uncertainty examples for the rest of the paper.

```{r}
#| eval: false
#| echo: false
# Get map data: do this once and save
my_map_data <- get_urbn_map("counties", sf = TRUE) |>
  filter(state_name=="Iowa")
save(my_map_data, file="data/iowa_map.rda")
```

```{r}
#| eval: false
#| echo: false
# Get centroid because rgeos is depreciated
load("data/iowa_map.rda")
centroids <- as_tibble(gCentroid(as(my_map_data$geometry, "Spatial"), byid = TRUE))
my_map_data$cent_long <- centroids$x
my_map_data$cent_lat <- centroids$y
save(my_map_data, file="data/my_map_data.rda")
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-data
#| fig-cap: "The first 5 observations of the data used for the spatial uncertainty examples along with the boundaries of each county. The map boundaries are the Iowa county boundaries, however the 'temperature' data is not representative of the average temperature in Iowa. The temperature and standard error represent the average of the daily high temperature and the standard error of that average respectively."
#| fig-subcap: 
#|   - "Data Table"
#|   - "Map Boundaries"
#| layout-nrow: 1
#| cap-location: "bottom"

# get data
load("data/my_map_data.rda")

# seed for sampling
set.seed(1997)

# data dimension for sampling
n <- dim(my_map_data)[1]

# Make palettes
longpal <- rev(sequential_hcl(13, palette = "YlOrRd"))
basecols <- longpal[3:10]
breaks <- 21:29 
breakslong <- 18:32 
names(basecols) <- seq(8)
names(longpal) <- -1:11

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The Noisy Work of Uncertainty Visualisation Research: A Review},
  pdfauthor={Harriet Mason; Dianne Cook; Sarah Goodwin; Emi Tanaka; Susan Vanderplus},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\date{October 18, 2024}
\title{\bf The Noisy Work of Uncertainty Visualisation Research: A
Review}
\author{
Harriet Mason\\
Department of Econometrics and Business Statistics, Monash University\\
and\\Dianne Cook\\
Department of Econometrics and Business Statistics, Monash University\\
and\\Sarah Goodwin\\
Department of Human Centred Computing, Monash University\\
and\\Emi Tanaka\\
Biological Data Science Institute, The Australian National University\\
and\\Susan Vanderplus\\
Statistics Department, University of Nebraska--Lincoln\\
}
\maketitle

\bigskip
\bigskip
\begin{abstract}
Uncertainty visualisation is quickly becomming a hot topic in
information visualisation. Exisiting reviews in the field take the
definition and purpose of an uncertainty visualisation to be self
evident which results in a large amout of conflicting information. This
conflict largely stems from a conflation between uncertainty
visualisations designed for decision making and those designed to
prevent false conclusions. We coin the term ``signal suppression'' to
describe a visualisation that is designed for preventing false
conclusions, as the approach demands that the signal (i.e.~the
collective take away of the estimates) is suppressed by the noise
(i.e.~the variance on those estimates). We argue that the current
standards in visualisation suggest that uncertainty visualisations
designed for decision making should not be considered uncertainty
visualisations at all. Therefore, future work should focus on signal
suppression. Effective signal suppression requires us to communicate the
signal and the noise as a single ``validity of signal'' variable, and
doing so proves to be difficult with current methods. We illustrate
current approaches to uncertainty visualisation by showing how they
would change the visual apprearance of a choropleth map. These maps
allow us to see why some methods succeed at signal suppression, while
others fall short. Evaluating visualisations on how well they perform
signal suppression also proves to be difficult, as it involves measuring
the effect of noise, a variable we typically try to ignore. We suggest
authors use qualitative studies or compare uncertainty visualisations to
the relevant hypothesis tests.
\end{abstract}

\newpage
\spacingset{1.9} % DON'T change the spacing!

\section{Introduction}\label{introduction}

@@

Among these summary tools, information visualisations are one of the
most powerful as they allow for quick and memorable communication that
allow us to identify quirks in our data that we didn't know to look for.
Datasets such as Anscombe's quartet \citep{anscombe} or the Datasaurus
Dozen \citep{datasaurpkg} show a case where visual statistics highlight
elements of the data that are invisible to the typical summary
statistics. Something as simple as sketching a distribution before
recalling statistics or making predictions can greatly increase the
accuracy of those measures \citep{Hullman2018, Goldstein2014}.

``Uncertainty visualisation'' is a relatively new field in research.
Early mentions of uncertainty visualisation start to appear in the late
1980s \citep{Ibrekk1987}, with geospatial information visualisation
literature from the early 1990s declaring this to be essential aspect of
any information display \citep{MacEachren1992, Carr1992}.
Figure~\ref{fig-ibrekk} depicts an example of the uncertainty
visualisations discussed in these early papers. Despite kicking off the
field, these papers did not define uncertainty visualisation. This has
led to a lack of consensus on what it means for a graphic to visualise
uncertainty, an issue we will return to later. Therefore, while the
field is considered to be quite new, many of the graphics used for
uncertainty visualisation have been around for much longer. For example,
box plots and histograms display variation which becomes synonymous with
uncertainty when we are using them to depict the variation of an
estimate. Today, there are an abundance of publications on the topic
which makes it timely to construct a review of the field. That is, now
that there is an overwhelming amount of information, it is valuable to
distil it into simple facts. In fact, there have already been several
reviews published but a central piece of discussion is missing.

\begin{figure}

\begin{minipage}{0.33\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-ibrekk-1.pdf}

}

\subcaption{\label{fig-ibrekk-1}Picture 1}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-ibrekk-2.pdf}

}

\subcaption{\label{fig-ibrekk-2}Picture 2}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-ibrekk-3.pdf}

}

\subcaption{\label{fig-ibrekk-3}Picture 3}

\end{minipage}%
\newline
\begin{minipage}{0.33\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-ibrekk-4.pdf}

}

\subcaption{\label{fig-ibrekk-4}Picture 4}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-ibrekk-5.pdf}

}

\subcaption{\label{fig-ibrekk-5}Picture 5}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-ibrekk-6.pdf}

}

\subcaption{\label{fig-ibrekk-6}Picture 6}

\end{minipage}%
\newline
\begin{minipage}{0.33\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-ibrekk-7.pdf}

}

\subcaption{\label{fig-ibrekk-7}Picture 7}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-ibrekk-8.pdf}

}

\subcaption{\label{fig-ibrekk-8}Picture 8}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-ibrekk-9.pdf}

}

\subcaption{\label{fig-ibrekk-9}Picture 9}

\end{minipage}%

\caption{\label{fig-ibrekk}A replication of the uncertainty
visualisations shown by \citet{Ibrekk1987} in one of the earliest
uncertainty visualisation experiments. Several visualisation methods
that are now unpopular (such as the pie chart) are used throughout this
paper.}

\end{figure}%

Reviews on uncertainty visualisation rarely offer tried and tested rules
for effective uncertainty visualisation, but rather, they comment on the
\emph{difficulties} faced when trying to summarise the field.
\citet{Kinkeldey2014} found most experimental methods to be ad hoc, with
no commonly agreed upon methodology, formalisations, or greater goal of
describing general principals. \citet{Hullman2016} noticed there is a
serious noise issue in the data coming from uncertainty visualisation
experiments. She commented on the prevalence of confounding variables
that make it unclear as to what exactly caused a subjects poor
performance on a set of particular questions. Mistakes due to
misunderstanding visualisations, misinterpreting questions, and
incorrectly applying heuristics are all combined into a single error
value. \citet{Spiegelhalter2017} commented that different plots are good
for different things, and disagreed with the goal of identifying a
universal best plot for all people and circumstances.
\citet{Griethe2006} did not identify common themes, but instead listed
the findings and opinions of a collection of papers.
\citet{uncertchap2022} summarised several cognitive effects that
repeatedly arise in uncertainty visualisation experiments, however these
effects were each discussed in isolation as a list of considerations an
author might make rather than an overarching theory of rules for
effective uncertainty visualisation. While these reviews are thorough in
scope, none discuss how the existing literature contribute to the
broader goal of uncertainty visualisation. The problem faced by the
literature is easily summarised with a famous quote by Henri Poincaré.

\begin{quote}
``Science is built up of facts, as a house is built of stones; but an
accumulation of facts is no more a science than a heap of stones is a
house.'' - Henri Poincaré (1905)
\end{quote}

@@

This review is broken into several parts that each reflect a different
approach to uncertainty visualisation. First, we look at graphics that
ignore uncertainty entirely and discuss why uncertainty should be
included at all. Second, we look at methods that consider uncertainty to
be just another variable and discuss the characteristics of uncertainty
that make it a unique visualisation problem. Third, we look at methods
that explicitly combine our estimate and its uncertainty and discuss the
limitations of these approaches. Fourth, we discuss methods that
implicitly include uncertainty by depicting a sample in place of an
estimate. Finally, we discuss how uncertainty visualisations can be
effectively evaluated. When discussing each of these methods, we will
repeatedly return to the \emph{purpose} of uncertainty visualisation and
the effectiveness of each approach in fulfilling that purpose.

\subsection{Spatial example}\label{spatial-example}

There are far too many uncertainty visualisations to exhaustively
discuss them all. Instead we focus on the changes made to a single plot,
the choropleth map. Due to the field's origins and focus in geospatial
information visualisation, there have been a large number of suggested
variations on the choropleth map that allow authors to include
uncertainty. Utilising a single example will help isolate the ideas we
are trying to convey. Therefore, it is important to remember that even
though we focus our discussion on the choropleth map, the theoretical
approach we outline in this review is useful to all uncertainty
visualisations regardless of their application. Additionally, our
examples focus on incorporating uncertainty through colour manipulation,
as that is the key visual channel used in a choropleth map. However, the
methods we discuss go beyond variations in a colour palette. Even though
they are not explicitly shown, visualisations that depict uncertainty
using layers such as position or shape, and more complicated graphics
that incorporate animation or interactivity are also within the scope of
this review. We will use the choropleth maps as a tool to clearly
highlight the costs and benefits of each approach.

Figure~\ref{fig-data} shows the first six rows and the geographical
boundaries of our data set. The temperature variable was generated using
a sine wave, that is
\(Temperature_i = 29-2\cdot{|Latitude_i - \sin{2 \cdot Longitude_i}|}\)
where the \(Longitude\) and \(Latitude\) are the longitude and latitude
of the county's centroid scaled to a standard normal distribution. Each
county's variance is independently, randomly sampled from a uniform
distribution. The low standard errors are drawn from a \(U_{[0,1]}\)
distribution, while the high standard errors are drawn from a
\(U_{[1,2]}\) distribution. As we are dealing with an average, the
sampling distribution would be approximately normal by the central limit
theorem, so each county temperature estimate is assumed to come from a
\(N(Temp_i, SE_{case,i})\). This is the data we will be using in our
spatial uncertainty examples for the rest of the paper.

\begin{figure}

\begin{minipage}{0.50\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-data-1.pdf}

}

\subcaption{\label{fig-data-1}Data Table}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-data-2.pdf}

}

\subcaption{\label{fig-data-2}Map Boundaries}

\end{minipage}%

\caption{\label{fig-data}The first 5 observations of the data used for
the spatial uncertainty examples along with the boundaries of each
county. The map boundaries are the Iowa county boundaries, however the
`temperature' data is not representative of the average temperature in
Iowa. The temperature and standard error represent the average of the
daily high temperature and the standard error of that average
respectively.}

\end{figure}%

\section{Ignoring uncertainty}\label{ignoring-uncertainty}

A good place to start might be at a deceptively straight forward
question, why should we include uncertainty at all?

\subsection{The choropleth map}\label{the-choropleth-map}

Figure~\ref{fig-choropleth} depicts a choropleth map of the counties of
Iowa. Each of these counties are coloured according to an estimate of
average daily temperature that was generated so that the values followed
a clear spatial trend. The variance of these estimates were simulated
such that the trend accounts for most of the variance in the plot in the
low variance case (so we should expect the trend to be visible) while in
the high variance, there is more variance within each county than
between all the counties, so we should expect it to (at least in some
capacity) overwhelm the spatial trend. Is this aspect of the data and
the spatial data it communicates clear in in the map? Is the strength of
the trend communicated through the visualisation?

\begin{figure}

\begin{minipage}{0.33\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-choropleth-1.pdf}

}

\subcaption{\label{fig-choropleth-1}Low Variance Data}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
